---
title: "protein_motif_database_searches_process_output"
author: "Janet Young\n"
date: "`r Sys.Date()`\n"
output: github_document
always_allow_html: true
---

# Goal 

Used protein motifs (output by MEME) to search selected genomes and proteomes - this script parses the output to help sort through the many weak hits we get.  

Search methods:  FIMO, MAST and hmmsearch.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(here)      # for tracking project home dir location
library(patchwork) # for showing >1 ggplot together
library(janitor)   # for clean_names() function 
library(universalmotif)
library(ggseqlogo)
library(Biostrings)
library(kableExtra)
library(xml2)       # to parse MAST xml output

source(here("motif_analysis/motif_analysis_functions.R"))

## load objects saved by protein_motif_logo_plots.Rmd
load(here("motif_analysis/protein_motif_logo_plots_Rscript_output/protein_motif_logo_plots_savedRobjects.Rdata"))

## MEME output base dir:
meme_dir <- here("motif_analysis/meme_protein_motifs/EZHIP_sequences_used_for_MEME_analyses.fa_meme/")
```



# Genomes and proteomes we're searching


We search 12 species, two databases for each, comprising three groups:

- `placental_other`: 'positive control' species - placental mammals where we should find EZHIP, including the KLP motif (motif 9)
- `afrotheria`: four Afrotheria species that seem to have lost KLP in EZHIP
- `marsupial/monotreme`: four outgroup species where we want to look for remote homologs 

For each species we search: 

- predicted proteomes downloaded from NCBI. Should work better for intron-containing genes, but may be incomplete for some genes
- 6-frame translations of the whole genome. More complete than the predicted proteome, but not as good for intron-containing genes, and may yield a lot of noise


```{r read 12 proteomes info}
#  Read info on the 12 proteome files:
selected_12proteomes_info <- here("motif_analysis/meme_protein_motifs/EZHIP_sequences_used_for_MEME_analyses.fa_meme/sequence_database_lists/selected_proteome_files.tsv") |> 
    read_tsv(show_col_types = FALSE) |> 
    mutate(genome=sapply(strsplit(output_tag, "_"), "[[", 1)) |> 
    relocate(genome)
```

```{r read all 12 proteome files to get seq descriptions}
selected_12proteomes_seqinfo <-  lapply(1:nrow(selected_12proteomes_info), function(i) {
    this_genome <- selected_12proteomes_info[i,"genome"] |> deframe()
    these_seqs <- selected_12proteomes_info[i,"proteome_file"] |> 
        deframe() |> 
        read_seq_file_JY(type="AA")
    these_seqs_info <- these_seqs[["info"]] |> 
        mutate(genome=this_genome) |> 
        relocate(genome)
    return(these_seqs_info)
}) |> 
    bind_rows()
```


```{r read 12 genomes information}
# Read info on the 12 genome-6frame files:
selected_12genomes6frames_info <- here("motif_analysis/meme_protein_motifs/EZHIP_sequences_used_for_MEME_analyses.fa_meme/sequence_database_lists/selected_genome_6frame_files.tsv") |> 
    read_tsv(show_col_types = FALSE) |> 
    mutate(genome=sapply(strsplit(output_tag, "_"), "[[", 1)) |> 
    relocate(genome)
```



Here are the 12 species we searched:

```{r}
selected_12genomes6frames_info |> 
    select(common_name, genome, group) |> 
    left_join(selected_12proteomes_info |> 
                  select(genome, proteome_file),
              by="genome") |> 
    mutate(proteome_file_short = sapply(strsplit(proteome_file, "/"), function(x) {x[length(x)]} )) |> 
    mutate(proteome_file_short = case_when(
        proteome_file_short=="protein.faa" ~ sapply(strsplit(proteome_file, "/"), function(x) {
            paste0(x[length(x)-1], "/", x[length(x)])
        } ), 
        TRUE ~ proteome_file_short
    )) |> 
    select(species=common_name, genome, species_group=group, proteome=proteome_file_short) |> 
    rename_with(~ str_replace_all(., "_", " ")) |>
    kable(format.args = list(big.mark = ","),
          digits=1) |> 
    kable_styling(full_width = FALSE)
```


# FIMO search output 

From [MEME suite documentation](https://meme-suite.org/meme/doc/fimo.html) - "FIMO scans a set of sequences for individual matches to each of the motifs you provide"

## FIMO search of 12 selected proteomes

```{r}
## get filenames 
fimo_proteome_files <- paste0(meme_dir, "meme.txt.FIMO.vs.selectedProteomes") |> 
    list.files(pattern="proteome", full.names = TRUE) |> 
    list.files(pattern="fimo.tsv", full.names = TRUE)


names(fimo_proteome_files) <- sapply(strsplit(fimo_proteome_files, "/"), function(x) { x[length(x)-1] })
names(fimo_proteome_files) <- sapply(strsplit(names(fimo_proteome_files), "\\."), function(x) { x[5] })
names(fimo_proteome_files) <- gsub("_proteome", "", names(fimo_proteome_files))
```


```{r read fimo proteome search output}
## read output
fimo_proteomes <- lapply(names(fimo_proteome_files), function(x) {
    fimo_proteome_files[[x]] |> 
        read_tsv(show_col_types = FALSE, comment="#") |> 
        mutate(proteome = x )
}) |> 
    bind_rows() |> 
    clean_names() |> 
    relocate(proteome) |> 
    ## add alternative motif ID
    left_join(EZHIP_motif_summary |> 
                  ## add species info
                  select(EZHIP_motif_id=motif_id, motif_alt_id=orig_name),
              by="motif_alt_id") |> 
    relocate(EZHIP_motif_id, .after=motif_alt_id) |> 
    left_join(selected_12proteomes_info |> 
                  select(genome, common_name, group),
              by=c("proteome"="genome")) |> 
    relocate(common_name, group, .after="proteome") |> 
    ## add info on the hit
    left_join(selected_12proteomes_seqinfo |> 
                  select(sequence_name=seq_id, seq_len, desc),
              by="sequence_name") |> 
    relocate(seq_len, desc, .after=sequence_name)
```


FIMO seems to be very lenient - there are many many weak hits for all motifs.

Therefore here we look only at motif 9 (the KLP motif) for FIMO. 

### FIMO motif 9 hits (KLP) in human proteome (top 5)

EZHIP itself is the only hit that looks good (EZHIP2 is not in the annotated proteome)

```{r}
fimo_proteomes |> 
    filter(EZHIP_motif_id=="EZHIP-9") |> 
    filter(proteome=="hg38") |> 
    slice_head(n=5) |>
    arrange(p_value) |> 
    show_fimo_match_alignment()
```



### FIMO motif 9 hits (KLP) in Afrotheria proteomes 

Show the best two hits by p-value for each genome - the alignments are unimpressive:

```{r}
fimo_proteomes |> 
    filter(EZHIP_motif_id=="EZHIP-9") |> 
    filter(group=="afrotheria") |> 
    ### per genome, choose best 2 hits
    group_by(proteome) |> 
    arrange(p_value) |> 
    slice_head(n=2) |>
    ungroup() |> 
    show_fimo_match_alignment()
```



### FIMO motif 9 hits (KLP) in outgroup proteomes (marsupial/monotreme)

Show the best two hits by p-value for each genome - the alignments are unimpressive:

```{r}
fimo_proteomes |> 
    filter(EZHIP_motif_id=="EZHIP-9") |> 
    filter(group %in% c("marsupial","monotreme")) |> 
    ### per genome, choose best 2 hits
    group_by(proteome) |> 
    arrange(p_value) |> 
    slice_head(n=2) |>
    ungroup() |> 
    show_fimo_match_alignment(include_proteome=TRUE)
```


## FIMO search of 12 selected genome-6frame translations


```{r}
## get filenames 
fimo_6frame_files <- paste0(meme_dir, "meme.txt.FIMO.vs.selectedGenomes_6frames") |> 
    list.files(pattern="6frame", full.names = TRUE) |> 
    list.files(pattern="fimo.tsv", full.names = TRUE)

names(fimo_6frame_files) <- sapply(strsplit(fimo_6frame_files, "/"), function(x) { x[length(x)-1] })
names(fimo_6frame_files) <- sapply(strsplit(names(fimo_6frame_files), "\\."), function(x) { x[5] })
names(fimo_6frame_files) <- gsub("_6frame", "", names(fimo_6frame_files))
```


```{r read fimo genome search output}
## read output
fimo_6frames <- lapply(names(fimo_6frame_files), function(x) {
    fimo_6frame_files[[x]] |> 
        read_tsv(show_col_types = FALSE, comment="#") |> 
        mutate(genome = x )
}) |> 
    bind_rows() |> 
    clean_names() |> 
    relocate(genome) |> 
    ## add alternative motif ID
    left_join(EZHIP_motif_summary |> 
                  ## add species info
                  select(EZHIP_motif_id=motif_id, motif_alt_id=orig_name),
              by="motif_alt_id") |> 
    relocate(EZHIP_motif_id, .after=motif_alt_id) |> 
    left_join(selected_12proteomes_info |> 
                  select(genome, common_name, group),
              by="genome") |> 
    relocate(common_name, group, .after="genome") |> 
    ## add info on the hit
    left_join(selected_12proteomes_seqinfo |> 
                  select(sequence_name=seq_id, seq_len, desc),
              by="sequence_name") |> 
    relocate(seq_len, desc, .after=sequence_name)
```


### FIMO motif 9 hits (KLP) in human 6-frame translation (top 5)

Top hit is EZHIP (chrX), second hit is EZHIP2 (chr5), and the rest look unimpressive

```{r}
fimo_6frames |> 
    filter(EZHIP_motif_id=="EZHIP-9") |> 
    filter(genome=="hg38") |> 
    slice_head(n=5) |>
    arrange(p_value) |> 
    show_fimo_match_alignment(include_genome=TRUE, include_description=FALSE)
```

### FIMO motif 9 hits (KLP) in afrotheria 6-frame translations (top 2 per species)

Show the best two hits by p-value for each genome - they seem unimpressive:

```{r}
fimo_6frames |> 
    filter(EZHIP_motif_id=="EZHIP-9") |> 
    filter(group=="afrotheria") |> 
    ### per genome, choose best 2 hits
    group_by(genome) |> 
    arrange(p_value) |> 
    slice_head(n=2) |>
    ungroup() |> 
    show_fimo_match_alignment(include_genome=TRUE, include_description = FALSE)
```

### FIMO motif 9 hits (KLP) in marsupials/monotreme 6-frame translations (top 2 per species)

Show the best two hits by p-value for each genome - they seem unimpressive:

```{r}
fimo_6frames |> 
    filter(EZHIP_motif_id=="EZHIP-9") |> 
    filter(group %in% c("marsupial","monotreme")) |> 
    ### per genome, choose best 2 hits
    group_by(genome) |> 
    arrange(p_value) |> 
    slice_head(n=2) |>
    ungroup() |> 
    show_fimo_match_alignment(include_genome=TRUE, include_description = FALSE)
```



# MAST search output 

From [MEME suite documentation](https://meme-suite.org/meme/doc/mast.html):  "MAST searches sequences for matches to a set of motifs, and sorts the sequences by the best combined match to all motifs"

## MAST search of 12 selected proteomes


```{r}
## get filenames 
mast_proteome_files <- paste0(meme_dir, "meme.txt.MAST.vs.selectedProteomes") |> 
    list.files(pattern="proteome", full.names = TRUE) |> 
    list.files(pattern="mast.xml", full.names = TRUE)

names(mast_proteome_files) <- sapply(strsplit(mast_proteome_files, "/"), function(x) { x[length(x)-1] })
names(mast_proteome_files) <- sapply(strsplit(names(mast_proteome_files), "\\."), function(x) { x[5] })
names(mast_proteome_files) <- gsub("_proteome", "", names(mast_proteome_files))
```


```{r read mast output proteomes}
## read output - one row per motif (rather than per hit sequence)
mast_proteomes <- lapply(names(mast_proteome_files), function(x) {
    mast_proteome_files[[x]] |> 
        parse_mast_xml() |> 
        mutate(proteome = x )
}) |> 
    bind_rows() |>
    clean_names() |>
    relocate(proteome) |>
    ### add alternative motif ID
    left_join(EZHIP_motif_summary |>
                  # add species info
                  select(ezhip_motif_index=motif_id, orig_name),
              by="orig_name") |>
    relocate(ezhip_motif_index, .after=orig_name) |> 
    ### get motif diag
    add_EZHIP_motif_diag() |>
    mutate(motif_diag_ezhip = str_remove_all(motif_diag_ezhip, "EZHIP-")) |> 
    ### add more genome info
    left_join(selected_12proteomes_info |>
                  select(genome, common_name, group),
              by=c("proteome"="genome")) |>
    relocate(common_name, group, .after="proteome") |>
    ### add info on the hit
    left_join(selected_12proteomes_seqinfo |>
                  select(sequence_name=seq_id, seq_len, desc),
              by="sequence_name") |>
    relocate(seq_len, desc, .after=sequence_name)

# mast_proteomes |> slice_head(n=6)
```


```{r}
## Get summary level matches - one row per matched sequence, rather than one row per motif
mast_proteomes_summary_level <- mast_proteomes  |> 
    select(proteome, common_name, group, sequence_name, sequence_description,
           target_e_value_orig,
           target_e_value, 
           num_motifs, num_diff_motifs, motif_diag_ezhip ) |> 
    unique() |> 
    arrange(proteome, target_e_value) 
```


### MAST hits in human proteome (top 2 by target e-value):

To help us understand MAST output and scoring, look at the top 2 hits in the human annotated proteome (proteome doesn't include EZHIP2).  The "motif diag ezhip" column shows all the motifs found in a single target sequence, in order in the sequence (numbered as in Figure 2).

```{r}
mast_proteomes_summary_level |> 
    filter(proteome=="hg38") |> 
    slice_head(n=2) |> 
    show_mast_hits_summary_level()
```

Show the actual motif matches for those two hits. Second hit is unimpressive (EZHIP2 is not in the annotated proteome)

```{r}
mast_proteomes |> 
    filter(proteome=="hg38") |> 
    arrange(target_e_value) |> 
    show_all_mast_motifs_per_target_seq(num_targets = 2)
```

### MAST hits in marsupial/monotreme proteomes (top 2 by target e-value, regardless of which motifs they contain):

```{r}
mast_proteomes_summary_level |> 
    filter(group %in% c("marsupial","monotreme")) |> 
    group_by(proteome) |> 
    arrange(target_e_value) |> 
    slice_head(n=2) |> 
    ungroup() |> 
    show_mast_hits_summary_level(include_proteome = TRUE)
```



Show best hits in outgroup proteomes, motif level (top 2 per genome by target p-value):

```{r, message=FALSE}
mast_proteomes |> 
    filter(group %in% c("marsupial","monotreme")) |> 
    group_by(proteome) |> 
    arrange(target_e_value) |> 
    show_all_mast_motifs_per_target_seq(num_targets = 2)
```





### MAST hits in marsupial/monotreme proteomes (top 2 by target e-value, only hits containing motif 9):

A lot of the best MAST hits don't contain a match to motif 9 (the KLP). Here the top hits if we restrict ourselves to only those that contain a motif 9 match (top 2 per proteome) - they are unimpressive (alignments are below the table):

```{r}
mast_filt_hits_temp <- mast_proteomes_summary_level |> 
    filter(group %in% c("marsupial","monotreme")) |> 
    filter(str_detect(motif_diag_ezhip, "9"))  |> 
    group_by(proteome) |> 
    arrange(target_e_value) |> 
    slice_head(n=2) |> 
    ungroup() 

mast_filt_hits_temp|> 
    show_mast_hits_summary_level(include_proteome = TRUE)
```




Show best hits in outgroup proteomes that contain motif 9, motif level (top 2 per genome by target p-value, only hits containing motif 9):

```{r, message=FALSE}
mast_proteomes |> 
    filter(group %in% c("marsupial","monotreme")) |>
    filter(sequence_name %in% mast_filt_hits_temp$sequence_name) |> 
    group_by(proteome) |>
    show_all_mast_motifs_per_target_seq(num_targets = 2)
```



## MAST search of 12 selected genome-6frame translations



```{r}
## get filenames 
mast_6frame_files <- paste0(meme_dir, "meme.txt.MAST.vs.selectedGenomes_6frames") |> 
    list.files(pattern="6frame", full.names = TRUE) |> 
    list.files(pattern="mast.xml", full.names = TRUE)

names(mast_6frame_files) <- sapply(strsplit(mast_6frame_files, "/"), function(x) { x[length(x)-1] })
names(mast_6frame_files) <- sapply(strsplit(names(mast_6frame_files), "\\."), function(x) { x[5] })
names(mast_6frame_files) <- gsub("_6frame", "", names(mast_6frame_files))
```



```{r read mast output on 12 genomes 6 frame translations}
## read output - one row per motif (rather than per hit sequence)
mast_6frames <- lapply(names(mast_6frame_files), function(x) {
    mast_6frame_files[[x]] |> 
        parse_mast_xml() |> 
        mutate(genome = x)
}) |> 
    bind_rows() |>
    clean_names() |>
    relocate(genome) |>
    ### add alternative motif ID
    left_join(EZHIP_motif_summary |>
                  # add species info
                  select(ezhip_motif_index=motif_id, orig_name),
              by="orig_name") |>
    relocate(ezhip_motif_index, .after=orig_name) |>
    ### get motif diag
    add_EZHIP_motif_diag() |>
    mutate(motif_diag_ezhip = str_remove_all(motif_diag_ezhip, "EZHIP-")) |>
    ### add more genome info
    left_join(selected_12genomes6frames_info |>
                  select(genome, common_name, group),
              by="genome") |>
    relocate(common_name, group, .after="genome") |>
    ### parse the orf name to get coordinates
    parse_getorf_seq_description(description_colname = "sequence_description",
                                 sequence_name_colname = "sequence_name") |>
    dplyr::rename(target_len_nt=width) |> 
    mutate(target_len_aa=target_len_nt / 3) |>
    relocate(sequence_description, chromosome, start, end, target_len_nt,  strand, .after=motif_diag_ezhip)  |>
    relocate(target_len_aa, .after=sequence_name)

# mast_6frames |> slice_head(n=6)
```



```{r}
## Get summary level matches - one row per matched sequence, rather than one row per motif
mast_6frames_summary_level <- mast_6frames  |> 
    select(genome, common_name, group, sequence_name, sequence_description,
           target_e_value_orig,
           target_e_value, 
           num_motifs, num_diff_motifs, motif_diag_ezhip,
           chromosome, start, end, strand) |> 
    unique() |> 
    arrange(genome, target_e_value) 
```



### MAST hits in human hg38 6-frame translations (top 5 by target e-value):

```{r}
mast_hits_temp <- mast_6frames_summary_level |> 
    filter(genome=="hg38") |> 
    slice_head(n=5)
mast_hits_temp |> 
    show_mast_hits_summary_level(include_description=FALSE,
                                 extra_cols=c("chromosome","start","end","strand"))
```


Show the actual motif matches for those five hits - alignments for hits beyond EZHIP and the two chr5 ORFs containing each end of EZHIP2 look unimpressive

```{r}
mast_6frames |> 
    filter(sequence_name %in% mast_hits_temp$sequence_name) |> 
    arrange(target_e_value) |> 
    show_all_mast_motifs_per_target_seq(#num_targets = 5,
        include_description=FALSE, 
        show_target_orf_coords=TRUE)
```


### MAST hits in marsupial/monotreme 6-frame translations (top 2 per genome by target e-value):

Note that these all involve only the repetitive motifs 5 and 6:

```{r, message=FALSE}
mast_temp_outgroup_hits <- mast_6frames_summary_level |> 
    filter(group %in% c("marsupial","monotreme")) |> 
    group_by(genome) |> 
    arrange(target_e_value) |> 
    slice_head(n=2) |> 
    ungroup() 
mast_temp_outgroup_hits |> 
    show_mast_hits_summary_level(include_description=FALSE,
                                 include_genome=TRUE,
                                 extra_cols=c("chromosome","start","end","strand"))
```



Show the actual motif matches for those hits:

```{r}
mast_6frames |> 
    filter(sequence_name %in% mast_temp_outgroup_hits$sequence_name) |> 
    arrange(genome, target_e_value) |> 
    # group_by(genome) |> 
    # arrange(target_e_value) |> 
    # ungroup() |> 
    show_all_mast_motifs_per_target_seq(include_genome=TRUE,
                                        include_description=FALSE,
                                        show_target_orf_coords=TRUE)
```




### MAST hits in marsupial/monotreme 6-frame translations that contain motif 9 (top 2 per genome by target e-value):

We'd be more impressed by a hit if it contained something that looks like motif 9, so filter to only show those. Here the motif 9 matches are weak, and matches to the repetitive motifs 5 and 6 bump up the MAST score.

```{r, message=FALSE}
mast_temp_outgroup_hits <- mast_6frames_summary_level |> 
    filter(group %in% c("marsupial","monotreme")) |> 
    filter(str_detect(motif_diag_ezhip, "9")) |> 
    group_by(genome) |> 
    arrange(target_e_value) |> 
    slice_head(n=2) |> 
    ungroup() 
mast_temp_outgroup_hits |> 
    show_mast_hits_summary_level(include_description=FALSE,
                                 include_genome=TRUE,
                                 extra_cols=c("chromosome","start","end","strand"))
```



Show the actual motif matches for those hits:

```{r}
mast_6frames |> 
    filter(sequence_name %in% mast_temp_outgroup_hits$sequence_name) |> 
    arrange(genome, target_e_value) |> 
    show_all_mast_motifs_per_target_seq(include_genome=TRUE,
                                        include_description=FALSE,
                                        show_target_orf_coords=TRUE)
```



# hmmsearch output


### hmmsearch of 12 selected proteomes


```{r read hmmsearch output 12 proteomes}
# orig motifs (includes trimmed version of motif 9)
meme_hmmsearch_12proteomes <- paste0(meme_dir, "motifs_hmmsearch_proteomes_max/") |>
    list.files(pattern="_proteome.domtblout.txt") |> 
    lapply(function(x) {
        if (grepl("all_ten_motifs", x)[1]) {
            this_genome <- gsub("_proteome","",strsplit(x, "\\.")[[1]][5])
        }
        if (grepl("trimmed", x)[1]) {
            this_genome <- gsub("_proteome","",strsplit(x, "\\.")[[1]][6])
        }
        if(!exists ("this_genome")) {
            stop("\n\nERROR - couldn't figure out how to get genome tag from filename\n\n")
        }
        # return(this_genome)
        # cat("reading file ",x,"\n")
        output <- paste0(meme_dir, "motifs_hmmsearch_proteomes_max/", x) |>
            readHmmerDomtbl(get_descriptions=TRUE) 
        if(is.null(output)) {return(NULL)}
        output <- output |>
            mutate(genome=this_genome) |>
            relocate(genome) 
        ### for the trimmed motif 9 we don't add ezhip_motif_index but for the rest we do
        if( grepl("trimmed",x)[1] ) {
            # cat ("   trimmed file!!\n")
            output <- output |> 
                mutate(ezhip_motif_index=NA_integer_)
        } else {
            output <- output |> 
                mutate(ezhip_motif_index =  as.integer(str_remove_all(query_name, "EZHIP-")))
        }
        return(output)
    }) |> 
    bind_rows()|> 
    left_join(selected_12proteomes_seqinfo |> 
                  select(seq_id, desc),
              by=c("target_name"="seq_id")) |>
    ## add common name and group
    left_join(selected_12proteomes_info |>
                  select(genome, common_name, group),
              by="genome") |>
    relocate(group, common_name) |> 
    ## rename a column
    dplyr::rename(target_len_aa = target_len)



# table(meme_hmmsearch_12proteomes$target_name %in% selected_12proteomes_seqinfo$seq_id)
#  TRUE 
# 2489

meme_hmmsearch_12proteomes_filt <- meme_hmmsearch_12proteomes  |>
    filter(domain_score >= 10)
# dim(meme_hmmsearch_12proteomes)
# [1] 2489   25
# dim(meme_hmmsearch_12proteomes_filt)
# [1] 203  25
```


```{r}
## hmmsearch default inclusion threshold is target_eval<= 0.01
meme_hmmsearch_12proteomes_filtInclusion <- meme_hmmsearch_12proteomes |> 
    ## page 140 of the HMM user guide says this should be how I can pick out things abnove inclusion threshold
    filter(target_eval <= 0.01 & domain_condEval <= 0.01)
```


```{r}
#### summary level for proteome
meme_hmmsearch_12proteomes_filt_summarized <- meme_hmmsearch_12proteomes_filt |> 
    # (for this I don't want to include trimmed version of motif 9)
    filter(str_detect(query_name, "trimmed", negate=TRUE)) |>
    ## collapse
    summarise_hmmsearch_results_each_ORF(include_coords_columns=FALSE) |> 
    ### add target seq description
    left_join(selected_12proteomes_seqinfo |>
                  select(target_name=seq_id, desc),
              by="target_name") |> 
    left_join(selected_12proteomes_info |>
                  select(genome, common_name, group),
              by="genome") 

# dim(meme_hmmsearch_12proteomes_filt_summarized)
# [1] 80 14
```




#### hmmsearch of 12 selected proteomes - use inclusion threshold filter


How many seqs in the proteome passed inclusion threshold per species? We'd expect to find EZHIP in horse (equCab3) but we don't - it's most likely just missing from the annotated proteome.


```{r}
meme_hmmsearch_12proteomes_filtInclusion |> 
    select(genome, target_name) |> 
    unique() |> 
    mutate(genome=factor(genome, levels=selected_12genomes6frames_info$genome)) |> 
    count(genome, .drop=FALSE) |> 
    arrange(desc(n)) |> 
    dplyr::rename(num_proteome_hits=n) |> 
    left_join(selected_12proteomes_info |> 
                  select(genome, common_name, group),
              by="genome") |> 
    relocate(num_proteome_hits) |> 
    rename_with(~ str_replace_all(., "_", " ")) |>
    kable() |> 
    kable_styling(full_width = FALSE)
```

Show accessions and descriptions for all hits that pass HMMER's inclusion threshold. The two hits in mouse (mm39) are just two splice isoforms of EZHIP

```{r}
meme_hmmsearch_12proteomes_filtInclusion |> 
    # filter(genome=="hg38") |>
    select(common_name, genome, target_name, description) |> 
    unique() |> 
    rename_with(~ str_replace_all(., "_", " ")) |>
    kable() |> 
    kable_styling(full_width = FALSE)
```

The only hit that's not already annotated as EZHIP is in platypus - the hit comprises two EZHIP-5 domain matches in one proline-rich sequence "basic salivary proline-rich protein 3-like". Show the hmmer scores here. We also use the target name (XP_007669498.2) to look at the match alignments in the [hmmsearch.txt output file](meme_protein_motifs/EZHIP_sequences_used_for_MEME_analyses.fa_meme/motifs_hmmsearch_proteomes_max/all_ten_motifs.hmm.hmmsearch.VS.mOrnAna1_6frame.txt) - they look unconvincing.

```{r}
meme_hmmsearch_12proteomes_filtInclusion |> 
    filter(genome=="mOrnAna1") |>
    arrange(query_name) |> 
    relocate(query_name, description) |> 
    select(-group, -common_name, -genome) |> 
    rename_with(~ str_replace_all(., "_", " ")) |>
    kable() |> 
    kable_styling(full_width = FALSE)
# filter(query_name=="EZHIP-8")
```



#### hmmsearch of 12 selected proteomes - without inclusion threshold filter

##### just human 

Just human - individual domain level. Show top 25 domain matches - domain scores drop off very quickly after the true EZHIP match.

```{r}
meme_hmmsearch_12proteomes |> 
    arrange(desc(domain_score)) |> 
    filter(genome=="hg38") |> 
    filter(str_detect(query_name, "trimmed", negate = TRUE)) |> 
    slice_head(n=25) |> 
    show_hmmsearch_hit_table(include_description=TRUE, 
                             include_genome=FALSE)
```



Now look at matches summarized by the target sequence (>1 motif match per target). I use a domain score threshold of >= 10. Sorting matches by total target domain score, showing top 2.  Second hit after EZHIP is not good.

```{r}
meme_hmmsearch_12proteomes_filt_summarized |> 
    filter(genome=="hg38") |> 
    arrange(desc(total_target_score)) |> 
    slice_head(n=2) |> 
    show_hmmsearch_summary_table(include_coords = FALSE,
                                 include_description=TRUE)
```



##### afrotheria

Afrotheria - not showing individual domain level hits for all domains, because we do have tblastn hits, and we're not asking about additional hits to anything other than motif 9 (see above for that)

Show matches summarized by the target sequence (>1 motif match per target), using a domain score threshold of >= 10. Sorting matches by total target domain score, showing top 2 for each species

mEleMax1 (Asiatic elephant) is the only genome for which EZHIP is present in the annotated proteome - that is a hit here but doesn't contain motif 9, as we know. 

The other genomes don't seem to have EZHIP annotated, and the only hits are weak, containing just one motif (see 'motif diag ezhip' column).

```{r}
meme_hmmsearch_12proteomes_filt_summarized |> 
    filter(group=="afrotheria") |> 
    group_by(genome) |> 
    arrange(desc(total_target_score)) |> 
    slice_head(n=2) |>
    ungroup() |> 
    show_hmmsearch_summary_table(include_coords = FALSE,
                                 include_description=TRUE)
```



##### marsupials/monotremes

Individual domain level. Show top 5 domain matches per species. We can use the 'target name' column to look up individual hits in the `*_proteome.txt` output files in [this folder](meme_protein_motifs/EZHIP_sequences_used_for_MEME_analyses.fa_meme/motifs_hmmsearch_proteomes_max) - these hits look unimpressive.

```{r}
meme_hmmsearch_12proteomes |> 
    filter(group %in% c("marsupial", "monotreme")) |> 
    filter(str_detect(query_name, "trimmed", negate = TRUE)) |> 
    group_by(genome) |> 
    arrange(desc(domain_score)) |> 
    slice_head(n=5) |> 
    ungroup() |> 
    show_hmmsearch_hit_table(include_description=TRUE, 
                             include_genome=FALSE)
```



Now look at matches summarized by the target sequence (>1 motif match per target). I use a domain score threshold of >= 10. Sorting matches by total target domain score, showing top 2 per species. Unimpressive.


```{r}
meme_hmmsearch_12proteomes_filt_summarized |> 
    filter(group %in% c("marsupial", "monotreme")) |> 
    group_by(genome) |> 
    arrange(desc(total_target_score)) |> 
    slice_head(n=2) |>
    ungroup() |> 
    show_hmmsearch_summary_table(include_coords = FALSE,
                                 include_description=TRUE)
```




### hmmsearch of 12 selected 6-frame translations

#### hmmsearch each motif - any species, using inclusion threshold


```{r read hmmsearch output 12 genomes 6frame}
# `meme_hmmsearch_12genomes_6frame` contains all matches found by hmmsearch (--max setting)
# `meme_hmmsearch_12genomes_6frame_filt` contains matches with domain score >=10

# orig motifs (includes trimmed version of motif 9)
meme_hmmsearch_12genomes_6frame <- paste0(meme_dir, "motifs_hmmsearch_genomes_max/") |>
    list.files(pattern="6frame.domtblout.txt") |> 
    lapply(function(x) {
        if (grepl("all_ten_motifs", x)[1]) {
            this_genome <- gsub("_6frame","",strsplit(x, "\\.")[[1]][5])
        }
        if (grepl("trimmed", x)[1]) {
            this_genome <- gsub("_6frame","",strsplit(x, "\\.")[[1]][6])
        }
        if(!exists ("this_genome")) {
            stop("\n\nERROR - couldn't figure out how to get genome tag from filename\n\n")
        }
        # cat("reading file ",x,"\n")
        output <- paste0(meme_dir, "motifs_hmmsearch_genomes_max/", x) |>
            readHmmerDomtbl(get_descriptions=TRUE) 
        if(is.null(output)) {return(NULL)}
        output <- output |>
            mutate(genome=this_genome) |>
            relocate(genome) 
        ### for the trimmed motif 9 we don't add ezhip_motif_index but for the rest we do
        if( grepl("trimmed",x)[1] ) {
            # cat ("   trimmed file!!\n")
            output <- output |> 
                mutate(ezhip_motif_index=NA_integer_)
        } else {
            output <- output |> 
                mutate(ezhip_motif_index =  as.integer(str_remove_all(query_name, "EZHIP-")))
        }
        
        output <- output |>
            parse_getorf_seq_description(description_colname = "description",
                                         sequence_name_colname = "target_name") |>
            dplyr::rename(target_len_aa=target_len) |>
            dplyr::rename(target_len_nt=width)
        return(output)
    }) |> 
    bind_rows() |> 
    left_join(selected_12genomes6frames_info |> 
                  select(genome, common_name, group), 
              by="genome") |> 
    relocate(group, common_name)
# meme_hmmsearch_12genomes_6frame
meme_hmmsearch_12genomes_6frame_filt <- meme_hmmsearch_12genomes_6frame  |>
    filter(domain_score >= 10)
# meme_hmmsearch_12genomes_6frame_filt
```


```{r}
## hmmsearch default inclusion threshold is target_eval<= 0.01
meme_hmmsearch_12genomes_6frame_filtInclusion <- meme_hmmsearch_12genomes_6frame |> 
    ## page 140 of the HMM user guide says this should be how I can pick out things abnove inclusion threshold
    filter(target_eval <= 0.01 & domain_condEval <= 0.01)
```

```{r}
# meme_hmmsearch_12genomes_6frame_filtInclusion |> 
#     filter(genome=="loxAfr3") |>
#     arrange(query_name) |> 
#     relocate(query_name) |> 
#     select(-group, -common_name, -genome, -chromosome, -start, -end, -strand)
# filter(query_name=="EZHIP-8")
```


#### hmmsearch - motif 9 hits 

##### hmmsearch - motif 9 hits in human

Human - requiring domain score >=10 we find a bunch of hits. Here I show the top 4. Notes on each:

- (1) EZHIP 
- (2) EZHIP2
- (3) a 134aa ORF called chr14_16745 with poor e-value (0.67). The motif 9 match is `ASSPSPPGR` (pos 8-16 of motif 9) - seems unimpressive
- (4) a 156aa ORF called chr10_107312 with poor e-value (1.1)

These hits are filtered on domain score >=10 - without that filter, there are a lot of even more unimpressive hits (e.g. a match to just `SSPSP` with domain score -1.5). This is a very sensitive search (too sensitive).

```{r}
meme_hmmsearch_12genomes_6frame |> 
    filter(genome=="hg38" & query_name=="EZHIP-9") |> 
    filter(domain_score >= 10) |>
    slice_head(n=4) |>
    show_hmmsearch_hit_table()
```

Show matches to motif 9 that pass HMMER's "inclusion threshold" - only EZHIP and EZHIP2 pass that threshold:

```{r}
meme_hmmsearch_12genomes_6frame_filtInclusion |> 
    filter(genome=="hg38" & query_name=="EZHIP-9") |> 
    show_hmmsearch_hit_table()
```

##### hmmsearch - motif 9 hits in afrotheria

Show best motif 9 matches in each of the four afrotheria genomes - best single motif 9 match in each genome (by target_eval). 

I show matches in tabular format here. I use the target name column to search the plain text hmmsearch.txt output file (e.g. [`all_ten_motifs.hmm.hmmsearch.VS.TriManLat1_6frame.txt`](meme_protein_motifs/EZHIP_sequences_used_for_MEME_analyses.fa_meme/motifs_hmmsearch_genomes_max/all_ten_motifs.hmm.hmmsearch.VS.TriManLat1_6frame.txt)) to look at the alignments between motif and target sequence. I have not written code to extract those alignments and display them here.

Here are the best motif 9 matches in those four afrotherian genomes - none have 'VRMR'.

The manatee (TriManLat1) match is the only one that passes HMMER's inclusion threshold. The rest have high e-values (>=0.29)

```
mEleMax1 - best motif9 hit is          SPSPPGELFPL
loxAfr3 - best motif9 hit is         ASSPSPPGLMFL
TriManLat1 - best motif9 hit is ATRHRASSPSPPG
mDugDug1 - best motif9 hit is          SPSPPGRLYLL
```

```{r}
meme_hmmsearch_12genomes_6frame |> 
    filter(group=="afrotheria" & query_name=="EZHIP-9") |> 
    # filter(domain_score >= 10)  |> 
    ## get best hit in each genome
    group_by(genome) |> 
    arrange(target_eval) |> 
    slice_head(n=1) |> 
    ungroup() |> 
    show_hmmsearch_hit_table()
# filter(target_eval < 0.1)
```


##### hmmsearch - motif 9 hits in marsupial/monotreme

Show best motif 9 match in each of the four marsupial/monotreme outgroup species genomes - when I look at the alignments none have 'VRMR'.

```
mMacEug1 - best motif9 hit is  PMQASSPSPPPSFFFFF
mMonDom1 - best motif9 hit is    RNSSPSPPRRHFVF
mSarHar1 - best motif9 hit is    RASSPAPPGKLFP
mOrnAna1 - best motif9 hit is   MRCSSPS
```

```{r}
meme_hmmsearch_12genomes_6frame |> 
    filter(group=="marsupial" | group=="monotreme") |> 
    filter(query_name=="EZHIP-9") |> 
    ## get best hit in each genome
    group_by(genome) |> 
    arrange(target_eval) |> 
    slice_head(n=1) |> 
    ungroup() |> 
    show_hmmsearch_hit_table()
```


#### hmmsearch all motifs combined - summarize matches per target ORF 


```{r}
#### summary level with for 6-frame translations
meme_hmmsearch_12genomes_6frame_filt_summarized <- meme_hmmsearch_12genomes_6frame_filt |> 
    # (for this I don't want to include trimmed version of motif 9)
    filter(str_detect(query_name, "trimmed", negate=TRUE)) |> 
    summarise_hmmsearch_results_each_ORF() |> 
    left_join(selected_12genomes6frames_info |> 
                  select(genome, common_name, group), 
              by="genome") |> 
    relocate(group, common_name)

# dim(meme_hmmsearch_12genomes_6frame_filt_summarized)
# [1] 3381   18
```

##### hmmsearch all motifs combined - human

I'll use human to figure out some reasonable filtering strategies at the level of summarized hits.

A reasonable filtering strategy seems to be: 

- ignore hits that only have the repetitive motifs 5/6/7 and no others
- require at least one motif with e-value < 0.1
- require at least 20aa worth of sequence aligned to any motif

Check that strategy in human - that gets me the ONLY 3 human regions we know about (chrX, plus the two very nearby bit on opposite strands on chr5).  The next-best thing after that was chr20_39436 - has two matches to motif 4 (total aln length = 24aa). When I look at the hit it's unimpressive.

```{r}
hmmsearch_filter_summarized_output <- function(x) {
    x |> 
        select(-group, -common_name) |> 
        ### sort by total target score
        arrange(desc(total_target_score)) |> 
        ### ignore hits that only have the repetitive motifs
        filter(motifs_present != "5") |> 
        filter(motifs_present != "6") |> 
        filter(motifs_present != "7") |> 
        # pairs
        filter(motifs_present != "5,6") |>
        filter(motifs_present != "5,7") |>
        filter(motifs_present != "6,7") |>
        # all three
        filter(motifs_present != "5,6,7") |>
        # need at least one motif with e-value < 0.1
        filter(best_target_eval < 0.1) |>
        ## need at least 20aa worth of sequence aligned to any motif
        filter(total_ali_width >= 20)   
}
```


```{r}
meme_hmmsearch_12genomes_6frame_filt_summarized |> 
    filter(genome=="hg38") |> 
    hmmsearch_filter_summarized_output() |> 
    # select(-genome) |> 
    show_hmmsearch_summary_table(include_genome=FALSE)
```

##### hmmsearch all motifs combined - Afrotheria

We're not checking Afrotherian genomes with motifs other than the KLP (motif 9) - we HAVE found an EZHIP gene or pseudogene in these species. The question we want to ask in these species is whether they have a KLP motif in any other gene.


```{r}
# meme_hmmsearch_12genomes_6frame_filt_summarized |> 
#     filter(group == "afrotheria") |> 
#     hmmsearch_filter_summarized_output() |> 
#     arrange(genome) |> 
#     show_hmmsearch_summary_table()
```


##### hmmsearch all motifs combined - marsupials/monotreme

Check matches in outgroup genomes (marsupials and monotreme) using the same filtering strategy - these are not convincing


```{r}
meme_hmmsearch_12genomes_6frame_filt_summarized |> 
    filter(group %in% c("marsupial", "monotreme")) |> 
    hmmsearch_filter_summarized_output() |> 
    show_hmmsearch_summary_table(include_genome=TRUE,
                                 include_coords=FALSE)
```



# Finished

```{r}
sessionInfo()
```
