---
title: "EZHIP_RNAseq_Example_Script_human"
output: github_document
always_allow_html: true
---

This code is an example script.  Here we process just the human Brawand tissue data - we used very similar code for the other datasets. 


Input = results of running bedtools multicov (RNA-seq read counts per gene)

Tasks:

- transform those counts to RPKMs
- generate a plot of expression levels of each gene in each tissue

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
```


Defining location of files of interest

```{r}
malik_h_dir <- "/fh/fast/malik_h/"

top_level_dir <- paste0(malik_h_dir, 
                        "grp/public_databases/NCBI/SRA/data/mammalian_expression_profiles/")

flagstats_dir <- paste0(top_level_dir, 
                        "human/human_Brawand/STAR_hg38_maxMultiHits1_renamed/")

dir.exists(flagstats_dir)
```

Obtaining filenames of flagstats files:

```{r}
flagstats_files <- list.files(flagstats_dir,
                              pattern="flagstats")
```


Define a function to read in flagstat files:

```{r}

readFlagstatsFile <- function (myFile, 
                               myDir=NULL) {
    if(!is.null(myDir)) {
        myFileFullPath <- paste(myDir, myFile, sep="/")
    } else {
        myFileFullPath <- myFile
        myDir <- getwd()
    }
    if(!file.exists(myFileFullPath)) {
        stop("ERROR! the file you specified does not exist. Maybe use the myDir option?")
    }
    
    dat <- scan(myFileFullPath, sep="\n", what="character", quiet=TRUE)
    
    dat <- strsplit(dat, " ")
    
    numReads <- as.numeric(sapply(dat,"[[",1))
    
    readCategory <- sapply( dat, function(x) {
        y <- x[4:length(x)]
        y <- paste(y, collapse=" ")
        return(y)
    })
    
    dat <- tibble(numReads=numReads,
                  description=readCategory,
                  file=myFile,
                  dir=myDir) |>
        mutate(file=gsub("_STAR\\.bam\\.flagstats", "", file)) 
    return(dat)
}

```


Read all flagstats files:

```{r}
brawand_flatstats_output <- lapply(flagstats_files, function(x) {
    flagstats_this_file <- readFlagstatsFile(x,  myDir=flagstats_dir) |> 
        
        filter(str_detect(description, "^mapped"))
}) |> 
    bind_rows()  |>
    select(-description)

brawand_flatstats_output |> 
    head(3)
```

Read bedtools multicov counts file: 

```{r}

counts_dir <- paste0(malik_h_dir,
                     "grp/malik_lab/Former_Lab_Members/PravruthaRaman/EZHIP_RNAseq/Counts/")



brawand_counts_file <- paste0(
    counts_dir,
    "Human_AncDup_hg38.bed.counts.human_Brawand.split.multicov")

brawand_counts <- brawand_counts_file |> 
    read_tsv(show_col_types = FALSE)

brawand_counts
```

Pivot longer, show first few rows

```{r}
brawand_counts_long <- brawand_counts |> 
    mutate(gene_width=end-start) |> 
    select(-c(chr, start, end)) |> 
    pivot_longer(cols = -c(name,gene_width),
                 names_to="RNAseq_sample",
                 values_to="count")
brawand_counts_long 
```

Add total number of reads sequenced for each sample (from flagstats files)

```{r}

brawand_counts_long <- left_join(brawand_counts_long, 
                                 brawand_flatstats_output |> 
                                     select(file, totReads=numReads), 
                                 by=c("RNAseq_sample"="file"))
brawand_counts_long |> 
    head(3)
```

Calculate reads per million sequenced, per kb of gene_width

```{r}

brawand_counts_long <- brawand_counts_long |> 
    mutate(rpkm = (count  /  (totReads/10^6) ) / (gene_width/10^3) )

brawand_counts_long |> 
    select(name, RNAseq_sample, count, rpkm)
```

Get sample name without replicate ID and sex, so we can plot those together

```{r}
brawand_counts_long <- brawand_counts_long |> 
    mutate(RNAseq_sample_noRep = gsub("\\d","",RNAseq_sample )) |> 
    mutate(RNAseq_sample_noRep = gsub("_Male","",RNAseq_sample_noRep )) |> 
    mutate(RNAseq_sample_noRep = gsub("_Female","",RNAseq_sample_noRep )) |> 
    relocate(RNAseq_sample_noRep, .after=RNAseq_sample)
brawand_counts_long
```

Now plot, as in Fig 4B

```{r, fig.height=3, fig.width=5}

EZHIP_color_scheme <- character()
EZHIP_color_scheme["EZHIP"] <- "goldenrod1"
EZHIP_color_scheme["EZHIP2"] <- "dodgerblue2"

p1 <- brawand_counts_long |> 
    ## tidy up the data a little more
    filter(!str_detect(RNAseq_sample_noRep, "human_Brain")) |> 
    select(tissue=RNAseq_sample_noRep, gene=name, rpkm) |> 
    mutate(tissue = tolower(str_remove_all(tissue, "human_"))) |> 
    mutate(gene = str_remove_all(gene, "Hum_")) |> 
    mutate(gene = str_replace_all(gene, "_Dup", "2")) |> 
    mutate(gene = str_remove_all(gene, "_Anc")) |> 
    ## plot
    ggplot(aes(x=tissue, y=rpkm,
               fill=gene, color=gene )) +
    geom_boxplot(whisker.linewidth=0.5, whisker.color="gray50",
                 median.linewidth = 0.5, median.colour="gray50") +
    geom_point(size=1.5, color ="gray50", position=position_dodge(width=0.75)) +
    labs(x="", y= "RPKM", title= "Human Brawand data") +
    theme_classic() +
    scale_fill_manual(values=EZHIP_color_scheme) +
    scale_color_manual(values=EZHIP_color_scheme) +
    expand_limits(y=10) +
    scale_y_continuous(breaks=c(0,5,10)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
p1
```



# Show R and package versions used

```{r}
sessionInfo()
```

