---
title: "protein_motif_logo_plots"
author: "Janet Young\n"
date: "`r Sys.Date()`\n"
output: github_document
always_allow_html: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)      # for tracking project home dir location
library(patchwork) # for showing >1 ggplot together
library(janitor)   # for clean_names() function 
library(universalmotif)
library(ggseqlogo)
library(Biostrings)

source(here("logo_plots/protein_motif_logo_plots_functions.R"))
```



# Read MEME output

Read meme.txt file using universalmotif::read_meme


```{r}
meme_dir <- here("logo_plots/meme_protein_motifs/EZHIP_sequences_used_for_MEME_analyses.fa_meme/")

meme_txt_file <- paste0(meme_dir, "meme.txt")

suppressMessages( meme_output <- universalmotif::read_meme(
    meme_txt_file, 
    readsites=TRUE, readsites.meta = TRUE ) )
# message = Could not find strand info, assuming +.
```


## Parse MEME output to get all instances


Put motif information into a simpler tbl called `all_motif_instances` that has one row for every motif instance.

```{r}
all_motif_instances <- lapply(names(meme_output[["sites.meta"]]), function(x) {
    meme_output[["sites.meta"]][[x]] |> 
        as_tibble() |> 
        mutate(motif_long_name=x) 
}) |> 
    bind_rows() |> 
    clean_names() |> 
    dplyr::rename(query_seq=sequence, 
                  instance_pval=pvalue, 
                  instance_seq=site) |> 
    relocate(motif_long_name)
```

Rename motifs based on their order in the human sequence (`Human_EZHIP`), so that new names are EZHIP-1 to EZHIP-10:

```{r}
human_motif_instances <- all_motif_instances |> 
    filter(query_seq=="Human_EZHIP") |> 
    arrange(position) |> 
    mutate(ezhip_motif_index=row_number()) |> 
    mutate(motif_id = paste0("EZHIP-", ezhip_motif_index)) |> 
    relocate(motif_id, ezhip_motif_index)
```

Get a simple tbl called `EZHIP_motif_summary` with one row for each of the ten motifs, and add on the EZHIP-1-style name

```{r}
EZHIP_motif_summary <- lapply(meme_output[["motifs"]], function(x) {
    output <- x |>
        as.data.frame()|>
        as_tibble() |> 
        mutate(eval_string = x@extrainfo) |> 
        mutate(width= ncol(x))
}) |> 
    bind_rows() |> 
    select(-family, -organism, -alphabet, -pval, -qval, -bkgsites, -strand) |> 
    dplyr::rename(motif_long_name=name,
                  orig_name=altname) |> 
    mutate(orig_index = as.integer(str_remove_all(orig_name, "MEME-"))) |> 
    left_join(human_motif_instances |>
                  select(motif_id, motif_long_name, ezhip_motif_index),
              by="motif_long_name") |> 
    relocate(motif_id, orig_name, orig_index, ezhip_motif_index) |> 
    arrange(ezhip_motif_index)
```



## Get PCMs for each motif, determine most common amino acids and their frequencies

Get each motif as a PCM (position count matrix)

```{r}
meme_output_pcms <- lapply(meme_output[["motifs"]], function(x) {
    ## colnames are initially the amino acids in the motif name/consensus, rather than positions
    output <- convert_type(x,"PCM")@motif 
    colnames(output) <- 1:dim(output)[2]
    return(output)
})
names(meme_output_pcms) <- sapply(meme_output[["motifs"]], function(x) {x@name})
```

Now for each motif, we find the most common amino acid at each position, and its frequency. 

```{r}
motifs_most_common_each_pos <- lapply(names(meme_output_pcms), function(x) {
    meme_output_pcms[[x]] |> 
        get_most_common_eachPos() |> 
        mutate(motif_long_name=x)
}) |> 
    bind_rows() |> 
    left_join(EZHIP_motif_summary |> 
                  select(motif_long_name, motif_id, ezhip_motif_index, orig_name),
              by="motif_long_name") |> 
    relocate(motif_long_name, motif_id, ezhip_motif_index, orig_name) |> 
    arrange(ezhip_motif_index, pos)|> 
    mutate(most_common_perc = 100*most_common_freq)
```


## Output fasta files of all instances for each motif

We use these to make HMMs to search genomes for additional homologs

First we add EZHIP-style names to `all_motif_instances` to create `all_motif_instances_plusInfo`.

```{r}
all_motif_instances_plusInfo <- all_motif_instances |> 
    left_join(human_motif_instances |> 
                  select(motif_long_name, motif_id, ezhip_motif_index), 
              by="motif_long_name") |> 
    relocate(motif_id, ezhip_motif_index) |> 
    ## join MEME_style ID back to all_motif_instances
    left_join(EZHIP_motif_summary |> 
                  select(motif_long_name, orig_name, orig_index),
              by="motif_long_name") |> 
    relocate(orig_name, orig_index)
```

Then we use `all_motif_instances_plusInfo` to export fasta files of the motif instances for each motif (they'll be aligned by default) so that I can make HMMs to scan the genome with.


```{r}
all_motif_instances_forFasta <- split(all_motif_instances_plusInfo, all_motif_instances_plusInfo$motif_id)

all_motif_instances_forFasta <- lapply(all_motif_instances_forFasta, function(x) {
    aa <- AAStringSet(x$instance_seq)
    names(aa) <- paste(x$motif_id, x$query_seq, sep="_")
    return(aa)
})

## actually write fasta files    
output_dir <- paste0(meme_dir, "motif_instance_fasta_files/")

temp <- lapply(names(all_motif_instances_forFasta), function(x) {
    outfile <- paste0(output_dir, x, ".fasta")
    writeXStringSet(all_motif_instances_forFasta[[x]], filepath = outfile)
    return(NULL)
})
rm(temp)
```


For motif 9 I will also save fasta file for a trimmed version, positions 3-16 - just the more conserved positions (AVRMRASSPSPPGR)

```{r}
motif9_trimmed <- narrow(all_motif_instances_forFasta[["EZHIP-9"]], start=3, end=16)
writeXStringSet(motif9_trimmed, filepath = paste0(output_dir, "EZHIP-9-trimmed_3to16.fasta"))
```



## Show motif logo plots, adding stars for conserved residues (80% threshold)


Show all logo plots - use * to highlight residues in >=80% of motif instances


```{r logo plots star80, fig.height=12, fig.width=5, warning=FALSE, message=FALSE}
new_logo_plots_ggseqlogo <- lapply(1:nrow(EZHIP_motif_summary), function(i) {
    ## so I can make sure they're all on the same scale:
    max_width <-  max(EZHIP_motif_summary$width)
    
    ## get info on THIS motif
    this_motif_info <- EZHIP_motif_summary |> 
        filter(ezhip_motif_index==i)
    
    ## get conserved positions
    temp_df <- motifs_most_common_each_pos |> 
        filter(motif_id==this_motif_info$motif_id) |> 
        filter(most_common_freq>=0.8) |> 
        select(pos, most_common, most_common_freq) |> 
        mutate(y_pos=4.4) |> 
        mutate(symbol="*")
    
    p1 <- convert_type(meme_output[["motifs"]][[this_motif_info$orig_index]],"PPM")["motif"] |> 
        ggseqlogo(col_scheme=weblogo_chemistry_color_scheme_ggseqlogo) +
        guides(fill = "none") 
    
    ## change axis limits
    suppressMessages( p1 <- p1 + 
                          coord_cartesian(xlim=c(1,max_width),
                                          ylim=c(0,4.8)))
    
    ## plot aesthetics
    p1 <- p1 +
        geom_text(data=temp_df, 
                  aes(x=pos, y=y_pos, label=symbol),
                  inherit.aes = FALSE,
                  size=5) + 
        labs(title=this_motif_info$motif_id,
             subtitle=paste0(  "E=", this_motif_info$eval_string, " ; ",
                               "n=", this_motif_info$nsites,
                               " (", this_motif_info$orig_name, ")")) +
        scale_y_continuous(breaks = 0:4) + # Manually specify y-axis ticks 
        theme(spacing=unit(5, "points"),
              axis.text.x=element_blank(), 
              ## y axis:
              axis.line.y=element_line(linewidth=0.25), 
              axis.ticks.y=element_line(linewidth=0.25),
              axis.text.y=element_text(size=6),
              axis.title.y=element_text(size=8),
              title=element_text(size=7),
              plot.subtitle=element_text(size=6)) 
    
    p1
})

names(new_logo_plots_ggseqlogo) <- EZHIP_motif_summary$motif_id

### show all logos:
p1 <- wrap_plots( new_logo_plots_ggseqlogo,
                  ncol = 1, 
                  byrow = FALSE ) +
    plot_annotation(title="EZHIP motifs", 
                    subtitle="* = residues present in >=80% motif instances")

p1

ggsave(p1,
       file=here("logo_plots/Rscript_output/motifs_star80.pdf"),
       height=12, width=5)
```




# Finished - show package versions used

```{r}
sessionInfo()
```
